### 2016-11-29 real/draft.tex
[draft] Reactivate references
[draft] Rename L->Q, start work on asymptotic approx

\DeclareMathOperator{\GammaDist}{Gamma}
\newcommand{\Kalpha}{{K_\alpha}}
\newcommand{\Kbeta}{{K_\beta}}
\newcommand{\npack}{n_p}
\newcommand{\lmax}{\ell_{\rm max}}
\newcommand{\Lmax}{L_{\rm max}}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\firstDeriv}[1]{\frac{\partial}{\partial #1}}
\newcommand{\secDeriv}[1]{\frac{\partial^2}{\partial #1^2}} % second partial derivative
\newcommand{\secPartial}[2]{\frac{\partial^2}{\partial #1 \, \partial #2}} % second partial derivative
\newcommand{\tardis}{TARDIS}

@@

We consider a computer simulation that, given input parameters
$\bmtheta$ and a random-number seed, produces a set of simulated data
$ S = \{ (\nu_i, \ell_i): i=1\dots\npack \} = (\bmnu, \bml)$
consisting of $\npack$ packets, each with a frequency and a
luminosity. We distinguish between a random variable $L$ and its
concrete realization $\ell$.  To learn about $\bmtheta$, we need to
compare to data from a telescope available as luminosities in
frequency bins, $D = \{ ( \nu_b, \ell_b): b=1 \dots B\}$. The natural
estimate from a single simulation is to add up the luminosities of all
packets in a bin to estimate $L_b$.  With a different seed but
identical $\bmtheta$, both the number of packets in the bin and their
luminosities are usually different. Our goal is to capture what is
known about $L$ based on a \emph{single} simulation without recourse
to expensive repetitions or asymptotic approximations; i.e., we seek
$P(L_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
analysis, one would proceed to plug in the observed value $\ell_b$ and
perform an analysis to learn about $\bmtheta$. But here we focus on
the specific problem of $P(L_b \cond \nu_b, \npack, \bmnu, \bml )$ and
hence omit $\bmtheta$ throughout.

@@

We consider the total number of packets $\npack$ fixed, for example
chosen by a user. In contrast, the number of packets in a bin, $n_b$
is a realization of a random variable $N_b$ which we assume to follow
a Poisson distribution. With this assumption, we neglect bin-by-bin
correlations and treat each bin independently which allows us to focus
on a single bin and to drop the subscript $b$ in the remainder. Hence
\begin{align}
  \lleq{caa}
  p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^N}{N!}
  \qquad N = 0,1,\ldots,\infty ,
\end{align}
where the dependency of the Poisson mean $\lambda>0$ on $\nu$ is of no
concern to us. Let $\bmL$ denote the collection of luminosities of
packets in a certain bin. The total luminosity in that bin is then
given by
\begin{align}
  \lleq{cab}
  L = \sum_{j=1}^{N} L_j\,.
\end{align}
We assume that each $L_j$ has a distribution depending on $\nu$ and
other parameters $\bmphi$, and that each packet is logically
independent. Since $N$ follows a Poisson distribution, $L$ is said to
follow a compound Poisson process. The joint distribution then is
\begin{align}
  \lleq{cac}
  (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(L_j\cond \bmphi, \nu) \; p(N\cond \lambda),
\end{align}
where we used that the Poisson term does not depend on $\nu$
explicitly. In general, the distribution of $L_j$ could be a of a
different functional form in each bin. For simplicity, we assume it is
identical but may depend on parameters that differ from bin to bin. In
practice, the distribution of packets usually does not change abruptly
with frequency. We therefore assume that any parameter needed to fully
specify $p(L_j)$ is given explicitly as a continuous function of $\nu$
and the unknown parameters $\bmphi$. This allows us to use a small
number of parameters to describe $p(L)$ across the entire frequency
range instead of having a few parameters per bin which could lead to
tens of thousands of parameters for the applications we have in
mind. So we determine $\bmphi$ from fitting all simulated packets
simultaneously whereas $\lambda$ is a per-bin quantity.

Our goal of predicting $L$ in some bin given the simulation output
$(\npack,\bmnu, \bml)$ is captured in the probability density
$p(L\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
integrated out. To simplify the treatment, we assume that the
frequency bins are narrow or equivalently that $p(L \cond \nu, \npack,
\bmnu, \bml)$ does not vary within a bin. This implies that every
sample $L_j$ in the bin follows the same distribution.  As the
reference frequency, we choose the bin center.

@@

Since $L$ is fully determined by $(N,\bmL)$, it does not depend on

@@

  p(L\cond N,\bmL,\nu,\npack,\bmnu, \bml)
  = p(L\cond N, \bmL)
  &= \delta(L - \textstyle\sum_{j=1}^N L_j),

@@

each $L_j$ is assumed to occur at $\nu$.  Using the law of total

@@

  p(L\cond \nu, \npack,\bmnu, \bml)
  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(L\cond N,\nu, \npack,\bmnu, \bml) \\

@@

  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(L\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)

@@

  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_{j=1}^N L_j)

@@

  p(L\cond \nu, \npack,\bmnu, \bml)

@@

  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(L - \textstyle\sum_j L_j)

@@

  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(L_j\cond \bmphi, \nu), \\

@@

the prior $p(\bmphi)$ and the likelihood $p(L_j\cond \bmphi, \nu)$, we
have a prediction for $L$ which takes into account all possible $N$,

@@

luminosity $P(L_j \cond \bmphi, \nu)$ should be such that the integral

@@

distribution of $L_j$, the distribution of $L = \sum_N L_j$ should be

@@

% Based on staring at the distribution of luminosities,
To apply \refeq{pmp}, we first need to choose $P(L_j \cond \bmphi,
\nu)$. A histogram of the luminosities from a \tardis{} run indicates
a strongly skewed distribution with a single mode that is close to but
not at the sharp end point $\lmax$. There is a long tail for small
$\ell$. These features can be accomodated by a $\GammaDist$
distribution with shape parameter $\alpha$ and rate
parameter $\beta$
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}\, , \; x > 0 .
\end{align}
if we transform the data as
\begin{align}
  \lleq{gmf}
  x_i \equiv 1- \frac{\ell_i}{\lmax} > 0
\end{align}
and similarly for $L \to X$ because $\ell$ is bounded from above and
$x$ is bounded from below.  We first take the largest luminosity of
all \tardis{} packets across the entire frequency range and increase
it slightly to avoid numerical difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_i \ell_i,
\end{align}
where $ \epsilon=10^{-6}$.  We assume the total number of samples
across all bins is large, hence the effect of considering $\lmax$ a
constant independent of the data is negligible. We checked that
leaving $\lmax$ as a free parameter in a fit gave equivalent results.

We infer $P(X \cond \nu, \npack, \bmnu, \bmx)$ and only in the end
transform back to a probability density in $L$ via
\begin{align}
  \lleq{gmg}
  p(L  \cond \nu,  \npack, \bmnu, \bml) = p(X  \cond \nu,  \npack, \bmnu, \bmx) \frac{1}{\lmax} \,.
\end{align}
To simplify the numerics, we also rescale the frequencies by a
constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  A major
numerical simplification arises because of the $L \to X$
transformation. Rewriting the master~\refeq{pmp} formally as
\begin{align}
  \lleq{gmh}
  p(L  \cond \nu,  \npack, \bmnu, \bml) = \sum_{N} p(N \cond \nu,  n) p(L \cond N,
\nu, \npack, \bmnu, \bml),
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single-packet luminosity is narrow
in the sense that the standard deviation relative to the mean is at
the percent level. This implies that $p(L \cond \nu, \npack,
\bmnu, \bml)$ has one mode from each $N$ and all modes are clearly
separated. The transformation $L \to x$ effects a strong overlap
between the distributions $p(X \cond N, \nu, \npack, \bmnu, \bmx)$
with similar $N$, so $p(X \cond \nu, \npack, \bmnu, \bmx)$ can be well
approximated by a unimodal distribution; see \reffig{hedgehog}.

\begin{figure}[ht]
  \centering
  % \includegraphics[width=0.48\textwidth]{hedgehog}
  \caption{Left: Prediction $p(L \cond  \nu,  \npack, \bmnu, \bml)$, right: $p(x \cond  \nu,  \npack, \bmnu, \bmx)$. I sum over $N$ in both cases. }
  \fred{I still need to pretty this up. I used the old binned approach
    with 50 samples in a bin from which I computed the mean and
    variance to plug into \refeq{gmr} but I computed the sum over $N$
    instead of the integral.}
\label{fig:hedgehog}
\end{figure}

@@

@@

\bibliographystyle{plain}
% \bibliography{references}

>>>>

\DeclareMathOperator{\Expect}{\mathbb{E}}
\newcommand{\expect}[1]{\Expect\left[#1\right]}
\DeclareMathOperator{\GammaDist}{Gamma}
\DeclareMathOperator{\GaussianDist}{\mathcal{N}}
\newcommand{\Kalpha}{{K_\alpha}}
\newcommand{\Kbeta}{{K_\beta}}
\newcommand{\npack}{n_p}
\newcommand{\lmax}{\ell_{\rm max}}
\newcommand{\Lmax}{L_{\rm max}}
\newcommand{\Lumtot}{Q}
\newcommand{\lumtot}{q}
\newcommand{\Lum}{L}
\newcommand{\lum}{\ell}
\newcommand{\rmdx}[1]{\mbox{d} #1 \,} % differential
\newcommand{\firstDeriv}[1]{\frac{\partial}{\partial #1}}
\newcommand{\secDeriv}[1]{\frac{\partial^2}{\partial #1^2}} % second partial derivative
\newcommand{\secPartial}[2]{\frac{\partial^2}{\partial #1 \, \partial #2}} % second partial derivative
\newcommand{\tardis}{TARDIS}

@@

We consider a computer simulation that, given input parameters
$\bmtheta$ and a random-number seed, produces a set of simulated data
$ S = \{ (\nu_i, \ell_i): i=1\dots\npack \} = (\bmnu, \bml)$
consisting of $\npack$ packets, each with a frequency and a
luminosity. We distinguish between a random variable $L$ and its
concrete realization $\ell$.  To learn about $\bmtheta$, we need to
compare to data from a telescope available as luminosities in
frequency bins, $D = \{ ( \nu_b, \ell_b): b=1 \dots B\}$. The natural
estimate from a single simulation is to add up the luminosities of all
packets in a bin to estimate $\Lumtot_b$.  With a different seed but
identical $\bmtheta$, both the number of packets in the bin and their
luminosities are usually different. Our goal is to capture what is
known about $\Lumtot$ based on a \emph{single} simulation without recourse
to expensive repetitions or asymptotic approximations; i.e., we seek
$P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$. In an actual physics
analysis, one would proceed to plug in the observed value $\ell_b$ and
perform an analysis to learn about $\bmtheta$. But here we focus on
the specific problem of $P(\Lumtot_b \cond \nu_b, \npack, \bmnu, \bml )$ and
hence omit $\bmtheta$ throughout.

@@

We consider the total number of packets $\npack$ fixed, for example
chosen by a user. In contrast, the number of packets in a bin, $n_b$
is a realization of a random variable $N_b$ which we assume to follow
a Poisson distribution. With this assumption, we neglect bin-by-bin
correlations and treat each bin independently which allows us to focus
on a single bin and to drop the subscript $b$ in the remainder. Hence
\begin{align}
  \lleq{caa}
  p(N\cond\lambda) &= \frac{e^{-\lambda}\lambda^N}{N!}
  \qquad N = 0,1,\ldots,\infty ,
\end{align}
where the dependency of the Poisson mean, $\lambda>0$, on $\nu$ is
implicit. Let $\bmL$ denote the collection of luminosities of packets
in a certain bin. The total luminosity in that bin is then given by
\begin{align}
  \lleq{cab}
  \Lumtot = \sum_{j=1}^{N} \Lum_j\,.
\end{align}
We assume that each luminosity $\Lum_j$, and we usually omit the subscript $j$ to refer to a single luminosity, has a distribution depending on $\nu$ and
other parameters $\bmphi$, and that each packet is logically
independent. Since $N$ follows a Poisson distribution, $\Lumtot$
follows a \emph{compound Poisson process}. The joint distribution then is
\begin{align}
  \lleq{cac}
  (N,\bmL) &\ \sim\ p(N,\bmL\cond \bmphi,\lambda, \nu)
  = p(\bmL\cond N,\bmphi, \nu)\; p(N\cond \lambda) = \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu) \; p(N\cond \lambda),
\end{align}
where we used that the Poisson term does not depend on $\nu$
explicitly. In general, the distribution of $\Lum_j$ could be a of a
different functional form in each bin. For simplicity, we assume it is
identical but may depend on parameters that differ from bin to bin. In
practice, the distribution of packets usually does not change abruptly
with frequency. We therefore assume that any parameter needed to fully
specify $p(\Lum_j)$ is given explicitly as a continuous function of $\nu$
and the unknown parameters $\bmphi$. This allows us to use a small
number of parameters to describe $p(\Lumtot)$ across the entire frequency
range instead of having a few parameters per bin which could lead to
tens of thousands of parameters for the applications we have in
mind. So we determine $\bmphi$ from fitting all simulated packets
simultaneously whereas $\lambda$ is a per-bin quantity.
\fred{Reformulate when asymptotic approx. included: it has thousands of parameters}

Our goal of predicting $\Lumtot$ in some bin given the simulation output
$(\npack,\bmnu, \bml)$ is captured in the probability density
$p(\Lumtot\cond \nu, \npack,\bmnu, \bml)$ with all relevant parameters
integrated out. To simplify the treatment, we assume that the
frequency bins are narrow or equivalently that $p(\Lumtot \cond \nu, \npack,
\bmnu, \bml)$ does not vary within a bin. This implies that every
sample $\Lum_j$ in the bin follows the same distribution.  As the
reference frequency, we choose the bin center.

@@

Since $\Lumtot$ is fully determined by $(N,\bmL)$, it does not depend on

@@

  p(\Lumtot\cond N,\bmL,\nu,\npack,\bmnu, \bml)
  = p(\Lumtot\cond N, \bmL)
  &= \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j),

@@

each $\Lum_j$ is assumed to occur at $\nu$.  Using the law of total

@@

  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)
  &= \sum_N p(N\cond \nu, \npack,\bmnu, \bml) \, p(\Lumtot\cond N,\nu, \npack,\bmnu, \bml) \\

@@

  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} p(\Lumtot\cond N,\bmL,\nu, \npack,\bmnu, \bml)\, p(\bmL\cond N,\nu, \npack,\bmnu, \bml)

@@

  &= \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_{j=1}^N \Lum_j)

@@

  p(\Lumtot\cond \nu, \npack,\bmnu, \bml)

@@

  \sum_N p(N\cond n)\,\int \rmdx{\bmL} \delta(\Lumtot - \textstyle\sum_j \Lum_j)

@@

  p(\bmL\cond \bmphi, N, \nu) &= \prod_{j=1}^N p(\Lum_j\cond \bmphi, \nu), \\

@@

the prior $p(\bmphi)$ and the likelihood $p(\Lum_j\cond \bmphi, \nu)$, we
have a prediction for $\Lumtot$ which takes into account all possible $N$,

@@

luminosity $P(\Lum_j \cond \bmphi, \nu)$ should be such that the integral

@@

distribution of $\Lum_j$, the distribution of $\Lumtot = \sum_N \Lum_j$ should be

@@

% Based on staring at the distribution of luminosities,
To apply \refeq{pmp}, we first need to choose $P(\Lum_j \cond \bmphi,
\nu)$. A histogram of the luminosities from a \tardis{} run indicates
a strongly skewed distribution with a single mode that is close to but
not at the sharp end point $\lmax$. There is a long tail for small
$\ell$. These features can be accomodated by a $\GammaDist$
distribution with shape parameter $\alpha$ and rate
parameter $\beta$
\begin{align}
  \lleq{gma}
  \GammaDist(x | \alpha, \beta) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x}\, , \; x > 0 .
\end{align}
if we transform the data as
\begin{align}
  \lleq{gmf}
  x_i \equiv 1- \frac{\ell_i}{\lmax} > 0
\end{align}
and similarly for $\Lumtot \to X$ because $\ell$ is bounded from above and
$x$ is bounded from below.  We first take the largest luminosity of
all \tardis{} packets across the entire frequency range and increase
it slightly to avoid numerical difficulties at the end point
\begin{align}
  \lleq{gme}
  \lmax \equiv (1+\epsilon) \times \max_i \ell_i,
\end{align}
where $ \epsilon=10^{-6}$.  We assume the total number of samples
across all bins is large, hence the effect of considering $\lmax$ a
constant independent of the data is negligible. We checked that
leaving $\lmax$ as a free parameter in a fit gave equivalent results.

We infer $P(X \cond \nu, \npack, \bmnu, \bmx)$ and only in the end
transform back to a probability density in $\Lumtot$ via
\begin{align}
  \lleq{gmg}
  p(\Lumtot  \cond \nu,  \npack, \bmnu, \bml) = p(X  \cond \nu,  \npack, \bmnu, \bmx) \frac{1}{\lmax} \,.
\end{align}
To simplify the numerics, we also rescale the frequencies by a
constant from $\mathcal{O}(10^{15})$ to $\mathcal{O}(1)$.  A major
numerical simplification arises because of the $\Lumtot \to X$
transformation. Rewriting the master~\refeq{pmp} formally as
\begin{align}
  \lleq{gmh}
  p(\Lumtot  \cond \nu,  \npack, \bmnu, \bml) = \sum_{N} p(N \cond \nu,  n) p(\Lumtot \cond N,
\nu, \npack, \bmnu, \bml),
\end{align}
we see that the right-hand side is a mixture density. In our
application, the distribution of a single-packet luminosity is narrow
in the sense that the standard deviation relative to the mean is at
the percent level. This implies that $p(\Lumtot \cond \nu, \npack,
\bmnu, \bml)$ has one mode from each $N$ and all modes are clearly
separated. The transformation $\Lumtot \to x$ effects a strong overlap
between the distributions $p(X \cond N, \nu, \npack, \bmnu, \bmx)$
with similar $N$, so $p(X \cond \nu, \npack, \bmnu, \bmx)$ can be well
approximated by a unimodal distribution; see \reffig{hedgehog}.

\begin{figure}[ht]
  \centering
  % \includegraphics[width=0.48\textwidth]{hedgehog}
  \caption{Left: Prediction $p(\Lumtot \cond  \nu,  \npack, \bmnu, \bml)$, right: $p(x \cond  \nu,  \npack, \bmnu, \bmx)$. I sum over $N$ in both cases. }
  \fred{I still need to pretty this up. I used the old binned approach
    with 50 samples in a bin from which I computed the mean and
    variance to plug into \refeq{gmr} but I computed the sum over $N$
    instead of the integral.}
\label{fig:hedgehog}
\end{figure}

@@

\section{Asymptotic approximation} \label{sec:asymptotic}

In the case that there are many packets in every bin, some
simplifications arise as the distribution of individual packets
$p(\Lum_j \cond \bmphi, \nu)$ becomes irrelevant in the asymptotic regime
$n \to \infty$, which implies $\lambda \to \infty$. In that case, the
compound Poisson distribution can be approximated by a Gaussian
distribution provided that the second moment of $\Lum_j, \expect{\Lum_j^2}$
is finite. Theorem 4.3.1 of \cite{bening2002generalized} states
that
% \begin{align}
%   \label{eq:asympt-conv}
%   P\left(\frac{L- \lambda \expect{\Lum_j}}{\sqrt{\lambda \expect{\Lum_j^2}}}  < z \right)  \Rightarrow \Phi(z) \mbox{ as } \lambda \to \infty,
% \end{align}
the asymptotic distribution of $\Lumtot$ when $\lambda \to \infty$, given
the first and second moment of $\Lum_j$, is
\begin{align}
  \label{eq:asympt-normal}
  P(\Lumtot \cond \bmphi, \nu) = \GaussianDist \left( \Lumtot \cond \lambda \expect{\Lum_j}, \lambda \expect{\Lum_j^2} \right).
\end{align}
That allows us to model $\Lumtot$ in each bin \emph{without} making concrete
assumptions about the functional form of $P(\Lum_j \cond \bmphi, \nu)$; i.e., we may now view $\bmphi$ as the collection of first and second moment in each bin
\begin{align}
  \label{eq:phi-bin-moments}
  \bmphi = \left(\expect{\Lum_j(\nu_b)}, \expect{\Lum_j^2(\nu_b)} \right)_{b=1}^{n_b}.
\end{align}
The price for this flexibility is that $\bmphi$ may now contain many
parameters, and each one is determined only from the samples in a
single bin.

@@

\bibliographystyle{plain}
\bibliography{references}
